Analyzing Changes in Canadian Grocery Prices

### 2024-11-21 paper/paper.qmd
Updated discussion section of paper

## Interpretation of Grocery Pricing Dynamics

This study investigates the primary factors influencing current grocery prices in Canada by examining the Project Hammer dataset. Our Bayesian regression model reveals that prior prices (`price_old`), vendor identity (`vendor_name`), and recent price differences (`price_difference`) play significant roles in determining current prices. The influence of previous prices indicates price continuity across time, suggesting that grocery items tend to exhibit relatively stable pricing patterns with minor adjustments. Vendor-specific effects highlight the role of retailer pricing strategies, with some vendors maintaining consistently higher or lower prices. These vendor variations may reflect differences in market positioning, operational costs, or competitive strategies, impacting price consistency across the grocery sector.

Additionally, the presence of both positive and negative values in the `price_difference` variable illustrates the prevalence of both price increases and decreases, underscoring the dynamic nature of grocery pricing in response to external market forces. External factors, such as supply chain disruptions, inflation, and seasonal demand fluctuations, likely contribute to the observed variability. This model highlights the complexity of pricing dynamics and the importance of historical pricing and vendor-specific factors in shaping current grocery prices across major Canadian retailers.

## Implications for Competition and Consumer Costs

The findings from Project Hammer have broader implications for both competition in the grocery sector and consumer expenses. Vendor-specific price differences suggest that some retailers adopt competitive pricing strategies to attract cost-sensitive consumers, while others may leverage brand loyalty or perceived quality to justify higher prices. In this context, price-sensitive consumers may benefit from exploring pricing variations across vendors to find the best deals, while retailers may face pressure to adjust their prices to remain competitive.

The stability of historical prices as a predictor of current prices also highlights potential concerns about price rigidity in certain product categories. This rigidity may limit the effectiveness of competition, as vendors might rely on historical price baselines rather than actively adjusting prices in response to market demand. Policymakers interested in promoting competition in the grocery sector could consider encouraging more frequent pricing updates or transparency in pricing strategies, particularly for staple items that heavily impact household budgets.

## Limitations of Data and Model

This analysis has several limitations that may impact the accuracy and comprehensiveness of our findings. First, the dataset is limited to major grocery chains, which may not capture price variations across smaller or regional stores that could offer different pricing structures. Additionally, the lack of granular product-level data limits our ability to analyze category-specific trends (e.g., dairy versus produce) or regional differences within the same vendor.

The model’s focus on historical prices, vendor identity, and price differences also excludes potential external factors that may influence pricing, such as seasonal demand, promotions, or supply chain constraints. Including additional variables, such as the frequency of price adjustments or regional economic indicators, could enhance the model’s predictive power and provide a more comprehensive view of grocery pricing dynamics.

## Future Research Direction

Future research could expand this work by incorporating product-specific attributes, seasonal factors, and promotional data to capture a broader range of influences on grocery prices. Analyzing the correlation between demographic factors (e.g., income levels) and price patterns across vendors could reveal socio-economic impacts on different communities.

Increasing the model’s granularity to examine price dynamics at the regional or product-category level would provide deeper insights into pricing variations within stores or across regions. Additionally, including online grocery prices or specialty stores could offer a more comprehensive view of the Canadian market.

Applying machine learning techniques could enhance the model’s adaptability to rapidly changing market conditions, enabling more precise forecasts, especially amid challenges like inflation and supply chain disruptions, benefiting policymakers and consumers in understanding and managing grocery costs.

>>>>

## Correlation vs. Causation
This analysis utilizes Bayesian regression model to identify factors correlated with current grocery prices, including prior prices (`price_old`), vendor identity (`vendor_name`), and recent price differences (`price_difference`). While the model shows that historical prices and vendor-specific effects are strongly associated with current pricing, it’s essential to interpret these relationships cautiously in terms of causation.

The correlation between previous and current prices likely reflects price continuity rather than a causal mechanism. For example, stable pricing over time may arise from vendor policies or industry norms rather than from any inherent property of the product or market structure. Similarly, vendor identity correlates with price differences, but this does not imply that being a particular vendor causes higher or lower prices. Instead, vendor-specific prices likely reflect underlying operational costs, competitive positioning, or branding strategies. Caution is warranted in inferring causation, as unobserved factors—such as cost structures or consumer loyalty—may influence both vendor identity and pricing patterns.

The presence of both price increases and decreases (reflected in `price_difference`) further highlights the dynamic nature of grocery pricing. While these changes are correlated with the final prices observed, it would be speculative to conclude that specific price fluctuations directly cause current price levels. External factors like inflation, seasonal demand, and supply chain disruptions may drive these observed correlations, underscoring the need for a cautious interpretation of causality.

## Missing Data

The Project Hammer dataset contains some limitations in data coverage, which may impact the completeness and accuracy of our findings. Specifically, the data does not include smaller or regional grocery stores, which may have distinct pricing strategies and could offer lower or more competitive prices in certain areas. The absence of these smaller players may lead to an incomplete picture of the overall grocery market, as price variations across different types of stores are not captured.

Additionally, missing data on certain product categories or regions within vendors limits the granularity of the analysis. For example, without category-specific data (e.g., dairy, produce, or packaged goods), it’s challenging to examine whether price dynamics vary significantly across product types. This lack of detail may obscure category-level trends, which could be particularly relevant for staple items that heavily influence household spending. Future data collection efforts could focus on capturing a wider range of store types and product details to provide a more representative dataset.

Finally, missing information on factors like seasonal influences, promotions, or regional economic conditions means that some relevant variables are absent from the model. These factors could help explain variations in grocery pricing, particularly for products sensitive to external factors (e.g., fresh produce affected by seasonality). Addressing these data gaps could significantly improve the model’s ability to capture the full range of influences on grocery prices.

## Sources of Bias

The Project Hammer dataset and the corresponding analysis may contain several sources of bias that could affect the interpretation of results. One potential source of bias stems from the focus on major grocery chains, which may not accurately represent pricing dynamics across smaller, independent stores. Since major vendors typically have broader operational reach and standardized pricing strategies, the findings may be biased toward the practices of these larger corporations, potentially overestimating price stability or ignoring competitive pricing seen in smaller markets.

Another possible bias comes from the data’s retrospective nature, focusing on historical pricing as a predictor of current prices. This approach may inherently favor price continuity and overlook the role of short-term market disruptions, promotions, or external shocks. For instance, a focus on historical prices could understate the impact of sudden events (like supply chain breakdowns or economic crises) that may cause sharp price changes. This form of selection bias could lead to an overly conservative estimate of price volatility in the model.

Furthermore, vendor-specific effects could introduce bias if certain vendors follow distinct pricing philosophies, such as premium or discount positioning. If these vendors have unique market positions that attract particular types of consumers, the model may reflect not only price dynamics but also underlying consumer segmentation. This brand-driven bias could distort interpretations, particularly if certain vendors consistently maintain higher or lower prices due to factors outside competitive dynamics (e.g., brand loyalty or perceived product quality).
